---
title: "Group 3: Potato data"
author: "Abby Bryson, Brianna Brown, Paulo Izquierdo, Sikta Das Adhikari, Wendy Leuenberger"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

## Set up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages 

```{r, message = FALSE, warning = FALSE}
# Load packages
library(tidyverse)  # General data wrangling
library(magrittr)   # Piping. %>% = "and then", %<>% saves over
library(readxl)     # Read in excel spreadsheets (.xlsx)
library(prospectr)  # For managing wavelengths, derivative
```

### ggplot settings

```{r PlotSettings}
tbw <- theme_bw(base_size = 18)
th <- theme(panel.grid = element_blank())
nol <- theme(legend.position = 'none')
xl <- xlab('Wavelength (nm)')
yl <- ylab('First derivative')
```

## Initial data management

### Read in data

```{r ReadData, message = FALSE}
# Folder containing data, mapped on Wendy's computer. 
# Would need to change for different people.
DataLocation <- 'C:\\Users\\Wendy\\OneDrive - Michigan State University\\CSS 844\\Module 2\\Potatoes_8_26_20'

# Read in the potato data
# RAUDPC = relative area under disease progress curve
# First four rows are metadata; don't read them in 
RAUDPC <- read_excel(paste0(DataLocation, '\\2020 Potato Late Blight Trial RAUDPC 20220224.xlsx'), 
                     skip = 4, na = '.')

# Read in hyperspectral data
# Big file - 153 rows x 1394 columns. Won't need all columns
spect <- read_csv(paste0(DataLocation, '\\plotwise_values_new.csv'))

# NM 
vnir <- read_excel(paste0(DataLocation, 
                          '\\VNIR\ Band\ Name\ and\ range.xlsx'),
                   col_names = c('Layer', 'nm', 'Units'))

## Combined data
# We will create this document in the following code
# Can use this code to read the formatted data into other R scripts
# AllData <- read_csv('CombinedPotatoData.csv')
```

### Take a look at data

```{r}
# Disease data
RAUDPC %>% head
# Hyperspectral data
spect %>% head
# Layer names -> wavelengths (nm)
vnir %>% head
```

Add a wavelength column to vnir that matches the other files. Make that column just have the number, not Layer_number

```{r vnir}
vnir %<>% 
  mutate(Wavelength = Layer %>% str_remove('Layer_'))
```


### Which columns contain the same information

Output not shown - a bunch of TRUE/FALSE to show if the columns line up. 

```{r, results = 'hide'}
colnames(spect)[1:20]
RAUDPC[1:6]

# Column names aren't the same
colnames(RAUDPC) %in% colnames(spect) 

# Which columns correspond
# Rep is similar
spect$Rep %in% RAUDPC$Rep

# Rows have same data
spect$Row %in% RAUDPC$ROW

# Lines have mostly the same data
spect$Line %in% RAUDPC$LINE
RAUDPC$LINE %in% spect$Line

# Blocks have mostly the same data
spect$Block %in% RAUDPC$BLOCK

# PlotNumber and PLOT seem like mostly the same data
spect$PlotNumber %in% RAUDPC$PLOT
RAUDPC$PLOT %in% spect$PlotNumber

# Tier and range appear to be the same
spect$Range %in% RAUDPC$TIER
```

### Prep data for join

```{r, results = 'hide'}
# Let's use Rep and Plot to join
# Need to make sure the information lines up
# Column names - join by PLOT
spect %<>% rename(PLOT = PlotNumber)

# Make the Line values line up
spect$PLOT[spect$Line == 'AtlanticSpreader'] <- 'Spreader'

# Make the Reps line up
spect$Rep[spect$Line == 'AtlanticSpreader' &
            spect$Row == 1] <- 1
spect$Rep[spect$Line == 'AtlanticSpreader' &
            spect$Row == 5] <- 2
spect$Rep[spect$Line == 'AtlanticSpreader' &
            spect$Row == 8] <- 3

# Take a look (results not shown)
spect %>% head
RAUDPC %>% head
```

### Join data

```{r join, message = FALSE}
# Join
AllData <- RAUDPC %>% 
  full_join(spect)
# Move rows so index columns are all together
AllData %<>% 
  relocate(c(fid, id, Row, Range, Block, Line, Pixels),
           .after = Rep)

# Make sure each plot has a unique identifier
AllData$UniqueValue <- paste(AllData$PLOT, AllData$id, sep = '_')

# Save as csv in case it's helpful
# write_csv(AllData, file = 'CombinedPotatoData.csv')
```

## Data processing for analyses

### Functions

`MakeLong()`: Change data to long format

`PlotSummary()`: Take mean of the replicate plots and organize for plotting

`Derivative()`: Take the first derivative of the the wavelength data

`PrepPCA()`: Subset the data to work with the PCA

`RunPCA()`: Fit the PCA using the prepared data

`Classify()`: Use a threshold to determine 1 (blight) or 0 (not blight) categories for a particular date.

```{r}
# Pull out values, change into long format

MakeLong <- function(Data = data, Value = value){
  
  NewData <- Data %>% 
    # Pull out the columns with early blight data (EB)
    # Pull out the columns with a number and the provided value
    # I.e. Value = _mean, pulls out 1_mean to 274_mean
    select(UniqueValue, PLOT, LINE, Rep, id, Pixels,
           starts_with('EB'), num_range(1:274, Value))
  
  # Change from wide to long format. This part only changes the 
  # wavelengths to long. We may need to also change EB data
  NewData %<>% 
    pivot_longer(ends_with(Value), names_to = 'Wavelength',
                 values_to = 'Value')
  
  # Remove the Value (i.e. _mean) from the wavelength column
  # Change to numeric for use as numeric/plotting
  NewData$Wavelength %<>% str_remove(Value)
  
  NewData %<>% 
    left_join(vnir[,c('Wavelength', 'nm')])
  
  return(NewData)
}


PlotSummary <- function(Data = data, Colname = colname){
  
  # Pull out just the data wanted for plotting
  # Might need to also pull out certain EB data
  # Take the mean of whichever value (mean among replicates)
  NewData <- Data %>% 
    group_by(UniqueValue, PLOT, id, Wavelength, nm) %>% 
    summarize(Name = mean(Value, na.rm = TRUE)) %>%
    # Summarize isn't good at variable column names
    # Change name below
    rename_at('Name', ~ Colname) %>% 
    arrange(Wavelength)
  
  return(NewData)
  
}


Derivative <- function(Data = data, Colname = colname){
  # Pull out just the wavelength numbers and values
  # Needs to be a dataframe (not tibble)
  NewData <- Data %>% 
    ungroup %>% 
    select(-PLOT, -id, -Wavelength) %>% 
    arrange(nm) %>% 
    pivot_wider(names_from = nm,
                values_from = all_of(Colname)) %>% 
    as.data.frame
  
  # Make the UniqueValues into rownames
  # Only numeric values in data are allowed for derivative functions
  rownames(NewData) <- NewData$UniqueValue
  NewData$UniqueValue <- NULL
  
  # Take the first derivative
  Deriv <- t(diff(t(NewData), differences = 1)) %>% 
    as.data.frame()
  
  # Put the unique values back into the dataset
  Deriv$UniqueValue <- rownames(Deriv)
  
  # Make into a long version for plotting/manipulation
  DerivLong <- Deriv %>% 
    pivot_longer(cols = -UniqueValue,
                 names_to = 'nm',
                 values_to = 'Derivative',
                 names_transform = list(nm = as.numeric)) %>% 
    arrange(nm)
  
  # Return the long form
  return(DerivLong)
}

# Prepare the data for a PCA using data of particular wavelengths
PrepPCA <- function(Data, Min_nm, Max_nm){
  NewData <- Data %>% 
    # Choose the wavelengths to include
    filter(nm > Min_nm & nm < Max_nm) %>% 
    # Pull out just the necessary columns
    select(UniqueValue, Value, nm) %>% 
    # Remove NAs
    filter(!is.na(Value)) %>% 
    # Change into wide format as a prerequisite to the matrix
    pivot_wider(values_from = Value, names_from = nm)
  # Return the wide form
  return(NewData)
}

# Run the PCA on the simplified data
RunPCA <- function(Data, IDcol){
  # Remove the non-numeric identifying column 
  # (only numbers allowed in the matrix)
  JustNumbers <- Data %>% 
    select(-IDcol)
  # Run the PCA
  PCA <- JustNumbers %>% 
    prcomp
  # Pull out the two most meaningful PCs
  # Add the identifying column back in
  PCA_Data <- data.frame(
    PC1 = PCA$x[,1],
    PC2 = PCA$x[,2],
    label = Data[,IDcol]
  )
  # Return the PCA + identifying information
  return(PCA_Data)
}

# Change continuous data into classified 0/1 data
Classify <- function(Data, OrigColumn, NewColumn, Threshold){
  # Make new object (safer than working with original)
  Newdata <- Data
  # Create new column with 1/0 data
  Newdata[,NewColumn] <- ifelse(Newdata[,OrigColumn] > Threshold, 
                                1, 0)
  # Return data with the added column
  return(Newdata)
}

```

### Make datasets

```{r Process, message = FALSE}
# Mean values
MeanLong <- MakeLong(AllData, '_mean')
MeanPlot <- PlotSummary(Data = MeanLong, Colname = 'MeanValue')
MeanDeriv <- Derivative(MeanPlot, Colname = 'MeanValue')
MeanPrep <- PrepPCA(Data = MeanLong, Min_nm = 611, Max_nm = 784)
MeanPCA <- RunPCA(Data = MeanPrep, IDcol = 'UniqueValue')

# Median values
MedianLong <- MakeLong(AllData, '_median')
MedianPlot <- PlotSummary(Data = MedianLong,
                          Colname = 'MedianValue')
MedianDeriv <- Derivative(MedianPlot, Colname = 'MedianValue')
```


```{r NotFunction, include = FALSE}
# Code to manipulate data not in a function
# 
# Mean <- AllData %>% 
#   select(UniqueValue, PLOT, LINE, Rep, id, Pixels, 
#          ends_with(c('_mean')))
# Median <- AllData %>% 
#   select(UniqueValue, PLOT, LINE, Rep, id, Pixels, 
#          ends_with(c('_median')))
# 
# # Change into longer data format for plotting/data manipulation
# Mean %<>% 
#   pivot_longer(ends_with('_mean'), names_to = 'Wavelength',
#                values_to = 'Value')
# 
# Median %<>% 
#   pivot_longer(ends_with('_median'), names_to = 'Wavelength',
#                values_to = 'Value')
# 
# # Remove the RGB value
# Mean %<>% 
#   filter(Wavelength != 'Red_mean',
#          Wavelength != 'Green_mean',
#          Wavelength != 'Blue_mean')
# Median %<>% 
#   filter(Wavelength != 'Red_median',
#          Wavelength != 'Green_median',
#          Wavelength != 'Blue_median')
# 
# 
# # Remove _mean
# Mean$Wavelength %<>% str_remove('_mean')
# Median$Wavelength %<>% str_remove('_median')
# 
# # Average of sample
# PlotData <- Mean %>% 
#   group_by(UniqueValue, PLOT, id, Wavelength) %>% 
#   summarize(MeanValue = mean(Value, na.rm = TRUE))
# PlotDataMedian <- Median %>% 
#   group_by(UniqueValue, PLOT, id, Wavelength) %>% 
#   summarize(MedianValue = mean(Value, na.rm = TRUE))
# 
# PlotData$Wavelength %<>% as.numeric
# PlotDataMedian$Wavelength %<>% as.numeric
# 
# PlotData %<>% arrange(Wavelength)
# PlotDataMedian %<>% arrange(Wavelength)
```

### Initial plots of wavelengths

Here's plots with 10 random and all of the plots. The 10 random plots will change every time the code is run.

```{r, warning = FALSE, message = FALSE}
Ten <- unique(AllData$UniqueValue) %>% sample(10)

# Plot
ggplot(MeanPlot %>% filter(UniqueValue %in% Ten),
       aes(x = nm, y = MeanValue, 
                     color = UniqueValue)) +
  geom_line() + th + tbw + xl + #nol +
  labs(y = 'Mean value', 
       title = 'Mean value of 10 random plots') 
  

ggplot(MedianPlot %>% filter(UniqueValue %in% Ten),
       aes(x = nm, y = MedianValue, 
                     color = UniqueValue)) +
  geom_line() + th + tbw + xl + 
  labs(y = 'Median value',
       title = 'Median value of 10 random plots')

# Mean all plots
ggplot(MeanPlot,
       aes(x = nm, y = MeanValue, 
                     color = UniqueValue)) +
  geom_line() + th + tbw + nol + xl +
  labs(y = 'Mean value',
       title = 'Mean value of all plots')


# Median all plots
ggplot(MedianPlot,
       aes(x = nm, y = MedianValue, 
                     color = UniqueValue)) +
  geom_line() + th + tbw + nol + xl +
  labs(y = 'Median value',
       title = 'Median value of all plots')

```

### Troubleshooting

Plots with low wavelength values

Problem solved with new data (3/27/2022), so this section isn't needed. Code is still in the document but isn't run.

```{r, include = FALSE}
# Just the ones with a low value at the 150-230
Odd <- MeanPlot %>%
  filter(Wavelength == '200' &
           MeanValue <= 0.25)
ggplot(MeanPlot %>% filter(PLOT %in% Odd$PLOT),
       aes(x = nm, y = MeanValue,
                     color = UniqueValue)) +
  geom_line() + th + tbw + xl + nol +
  labs(y = 'Mean value',
       title = 'Mean values - questionable plots')
ggplot(MeanPlot %>% filter(PLOT %in% Odd$PLOT),
       aes(x = nm, y = MeanValue,
                     color = UniqueValue)) +
  geom_line() + th + tbw + xl + #nol +
  labs(y = 'Mean value',
       title = 'Mean values - questionable plots')

OddMedian <- MedianPlot %>%
  filter(Wavelength == '200' &
           MedianValue <= 0.25)
ggplot(MedianPlot %>% filter(PLOT %in% OddMedian$PLOT),
       aes(x = nm, y = MedianValue,
                     color = UniqueValue)) +
  geom_line() + th + tbw + nol +
  labs(y = 'Median value',
       title = 'Median values - questionable plots')
ggplot(MedianPlot %>% filter(PLOT %in% OddMedian$PLOT),
       aes(x = nm, y = MedianValue,
                     color = UniqueValue)) +
  geom_line() + th + tbw + #nol +
  labs(y = 'Median value',
       title = 'Median value - questionable plots')

```


### Derivatives


```{r, include = FALSE}
# Derivative code not in a loop
# 
# PlotData <- MeanPlot
# Matrix <- PlotData %>%
#   ungroup %>%
#   select(-PLOT, -id, -Wavelength) %>% 
#   arrange(nm)
# 
# Matrix %<>%
#   pivot_wider(names_from = nm,
#               values_from = MeanValue) %>%
#   as.data.frame()
# rownames(Matrix) <- Matrix$UniqueValue
# Matrix$UniqueValue <- NULL
# 
# d1_rl <- t(diff(t(Matrix), differences = 1))
# d1_rl %<>% as.data.frame()
# d1_rl$UniqueValue <- rownames(d1_rl)
# d1_rl_long <-
#   d1_rl %>%
#   pivot_longer(cols = -UniqueValue, names_to = 'nm',
#                values_to = 'Derivative')
# d1_rl_long$nm %<>% as.numeric
# d1_rl_long %<>% arrange(nm)
```

```{r DerivPlots, warning = FALSE}
# ggplot(d1_rl_long %>% filter(UniqueValue %in% Ten[1]),
#        aes(x = nm, y = Derivative, color = UniqueValue)) +
#   geom_line() + th + tbw + nol

ggplot(MeanDeriv %>% filter(UniqueValue %in% Ten),
       aes(x = nm, y = Derivative, color = UniqueValue)) +
  geom_line() + th + tbw + nol + yl + xl +
  ggtitle('First derivative of mean values at 10 random plots')

ggplot(MeanDeriv,
       aes(x = nm, y = Derivative, color = UniqueValue)) +
  geom_line() + th + tbw + nol + yl + xl +
  ggtitle('First derivative of mean values')


```

## PCA

```{r, include = FALSE}
# Test <- PrepPCA(MeanLong, 800, 900)
# TestRun <- RunPCA(Test, IDcol = 'UniqueValue')
# 
# MeanForPCA <- MeanLong %>% 
#   filter(nm > 800 & nm < 900) %>% 
#   select(UniqueValue, PLOT, Value, nm) %>% 
#   filter(!is.na(Value)) %>% 
#   mutate(Spreader = ifelse(PLOT == 'Spreader', 
#                            'Spreader', 'Not Spreader')) %>% 
#   pivot_wider(values_from = Value, names_from = nm)
# MeanPCA <- MeanForPCA %>% 
#   select(-UniqueValue, -PLOT, -Spreader) %>% 
#   prcomp
# PCA_Mean <- data.frame(
#   PC1 = MeanPCA$x[,1],
#   PC2 = MeanPCA$x[,2],
#   label = MeanForPCA$UniqueValue,
#   classification = MeanForPCA$Spreader
# )
# 
# ggplot(PCA_Mean, aes(x = PC1, y = PC2, 
#                      label = label, 
#                      col = classification)) +
#   geom_point() 
# 
# 
#   

```

### Variation explained by PCA
```{r}
JustNumbers <- MeanPrep %>%
  select(-UniqueValue)
  # Run the PCA
PCA <- JustNumbers %>% 
  prcomp

summary(PCA)$importance[1:3,1:5]
```


```{r}
ggplot(MeanPCA, aes(x = PC1, y = PC2)) +
  geom_point() 


```


## Classify EB

```{r}
# Note: Threshold is >, not >=

# Quantiles for choosing a threshold
# quantile(AllData$`EB%_dpi_0`, na.rm = TRUE)
# quantile(AllData$`EB%_dpi_23`, na.rm = TRUE)
# quantile(AllData$`EB%_dpi_26`, na.rm = TRUE)
# quantile(AllData$`EB%_dpi_30`, na.rm = TRUE)
# quantile(AllData$`EB%_dpi_37`, na.rm = TRUE)
# quantile(AllData$`EB%_dpi_43`, na.rm = TRUE)
quantile(AllData$`EB%_dpi_47`, na.rm = TRUE)

Mean23 <- Classify(MeanLong, 'EB%_dpi_23', 'EB_23', 10)
Mean26 <- Classify(MeanLong, 'EB%_dpi_26', 'EB_26', 10)

# Mean26 %>% 
#   select(`EB%_dpi_26`, EB_26) %>% 
#   unique

Mean47 <- Classify(AllData, OrigColumn = 'EB%_dpi_47', 
                   NewColumn = 'EB_47', Threshold = 24)
Mean47 %>% 
  select(`EB%_dpi_47`, EB_47) %>% 
  unique

```

## Logistic regression

```{r, message = FALSE}
# Add the PC1 and PC2 values to the data frame
Mean47 %<>% left_join(MeanPCA)

Accuracy <- vector(mode = 'numeric', length = 1000)

for(tt in 1:1000){

  # Rownames for the 80% of training data
  Include80 <- sample(x = length(Mean47$EB_47), 
                      size = 0.8 * length(Mean47$EB_47))
  Mean47_Train <- Mean47[Include80,]
  Mean47_Test <- Mean47[-Include80,]

  # 80/20
  MeanLogistic <- glm(EB_47 ~ PC1 + PC2, 
                      family = 'binomial',
                      data = Mean47_Train)
  summary(MeanLogistic)
  
  Mean47_Test %<>% 
    select(EB_47, PC1, PC2)

  Mean47_Test$Pred <- predict(MeanLogistic, 
                              newdata = Mean47_Test,
                              type = 'response')

  Mean47_Test$Pred01 <- ifelse(Mean47_Test$Pred >= 0.5, 1, 0)

  Correct <- sum(Mean47_Test$EB_47 == Mean47_Test$Pred01, 
                 na.rm = TRUE)
  Total <- Mean47_Test %>% filter(!is.na(EB_47)) %>% dim
  # print(paste(Correct, 'correct out of', Total[1]))
  Accuracy[tt] <- Correct/Total[1]
}

# Note - these values will change each time the markdown is run. 
# The means and medians should stay about the same.
Accuracy %>% summary
```

```{r}
AccuracyDF <- tibble(Accuracy = Accuracy,
                     Index = 1:length(Accuracy))
ggplot(AccuracyDF, aes(y = Accuracy, x = Index)) + 
  geom_point() +
  geom_hline(aes(yintercept = mean(Accuracy)), color = 'red') +
  tbw + th 

```


## Poisson regression

Similar relationship. We didn't include this model in our final presentation.

```{r, message = FALSE}
MeanPoisson <- glm(`EB%_dpi_47` ~ PC1 + PC2,
                   family = 'poisson',
                   data = Mean47)
summary(MeanPoisson)
plot(MeanPoisson)
```

## Make PCA plot with classified data

```{r, warning = FALSE}
# Classified version
ggplot(Mean47, aes(x = PC1, y = PC2, color = factor(EB_47))) +
  geom_point() +
  labs(color = 'Early blight day 47') +
  tbw + th

# Continuous version
# NA = gray
ggplot(Mean47, aes(x = PC1, y = PC2, color = `EB%_dpi_47`)) +
  geom_point() +
  labs(color = 'Early blight day 47')

```


<!-- ## Can we use python? -->

```{python, include = FALSE}
# Python code
print('Yes we can!')
```

